---
title: "RECOVER PCORnet Adult EHR Hub – Pulmonology Manuscript Display Items"
author: "Colby Lewis V" 
pi: "RECOVER; Beth Nauman "
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  wcmtheme::wcm_html: 
    toc: true
    toc_float: true
---


```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(dplyr)
library(tidyr)
library(lubridate)
library(readxl)
library(gtsummary)

library(survival)
library(survminer)
library(cards)

library(UpSetR)
library(boot)
```




## Load Data


```{r, include=F, eval=T}
# Load full CP
raw_cp_df_inc = read.csv("...", check.names=T)
raw_cp_df_inc$full_patid = paste(raw_cp_df_inc$site, "_", raw_cp_df_inc$patid, sep="")

print("loaded participants")
print(nrow(raw_cp_df_inc))
print("sites")
print(raw_cp_df_inc$site %>% unique())
```


## Tables for Incidence

### Category defn

```{r, include=F, eval=T}
raw_cp_df_inc = raw_cp_df_inc %>% #make sure out site dont slip in
  filter(flag_baseline_pulmonary == 0)

raw_cp_df_inc = raw_cp_df_inc %>% mutate(
  ymd_index = ymd(index.date),
  year_index = case_when(
    year(ymd_index) %in% c(2020, 2021, 2022, 2023) ~ year(ymd_index)
  ),
  index_date = as.Date(index.date, format = "%Y-%m-%d")
) %>% filter(
  ymd_index <= as.Date("2023-02-01"),
  !is.na(year_index),
  age >= 20,
  index_date >= as.Date("2020-03-01"),
  !is.na(dob)
)

cp_df = raw_cp_df_inc %>%
  mutate(year_index == factor(year_index))

cp_df = cp_df %>%
  mutate(clv_age = case_when(
    age < 40 ~ "20-39 years",
    age < 55 ~ "40-54 years",
    age < 65 ~ "55-64 years",
    age < 75 ~ "65-74 years",
    age < 85 ~ "75-84 years",
    age >= 85 ~ "85+ years"
  ))

cp_df = cp_df %>%
  mutate(clv_sex = case_when(
    Female == 1 ~ "Female",
    Male == 1 ~ "Male",
    T ~ "Other/Missing"
  ))

cp_df = cp_df %>%
  mutate(clv_race = case_when(
    Asian == 1 ~ "Asian",
    `Black.or.African.American` == 1 ~ "Black or African American",
    `White` == 1 ~ "White",
    `Other` == 1 ~ "Other",
    `Missing` == 1 ~ "Missing"
  ))

cp_df = cp_df %>%
  mutate(clv_hisp = case_when(
    `Hispanic..Yes` == 1 ~ "Hispanic: Yes",
    `Hispanic..No` == 1 ~ "Hipanic: No",
    `Hispanic..Other.Missing` == 1 ~ "Hispanic: Other/Missing"
  ))

cp_df = cp_df %>%
  mutate(out_bins = case_when(
    outpatient.no. == 0 ~ "0",
    outpatient.no. == 1 ~ "1",
    outpatient.no. < 6 ~ "2-5",
    outpatient.no. < 21 ~ "6-20",
    outpatient.no. > 20 ~ ">20"
  ))

cp_df = cp_df %>%
  mutate(em_bin = case_when(
    is.na(`emergency.visits.no.`) ~ "Missing",
    `emergency.visits.no.` > 1 ~ "2 or more",
    `emergency.visits.no.` == 1 ~ "1",
    `emergency.visits.no.` == 0 ~ "0"
  ))

cp_df = cp_df %>%
  mutate(bmi_clv = case_when(
    is.na(bmi) ~ "Missing",
    bmi < 18.5  ~ "Underweight (<18.5)",
    bmi < 25 ~ "Normal range (18.5 to <25)",
    bmi < 30 ~ "Overweight (25 to <30)",
    bmi >= 30  ~ "Obese (>30)"
  ))
```

### Excess burden bootstrap

#### combined:

```{r, message=FALSE, warning=FALSE, echo=FALSE}
burden_func_fast = function(data, indices, covid_bool) {
  sample_data = data[indices, ]
  percent = mean(sample_data$flag_pulmonary == 1) * 100
  
  return(percent)
}


burden_df = cp_df %>%
  select(hospitalized, covid, flag_pulmonary)

###################################################
this_df = burden_df %>% filter(covid==0)
results_0 = boot::boot(data=this_df,
                      statistic=burden_func_fast,
                      R=1000)
print(boot.ci(results_0, type="basic"))
###################################################
this_df = burden_df %>% filter(covid==1)
results_1 = boot::boot(data=this_df,
                      statistic=burden_func_fast,
                      R=5000)
print(boot.ci(results_1, type="basic"))

```

#### hosp:
```{r, message=FALSE, warning=FALSE, echo=FALSE}
hosp_df = burden_df %>%
  filter(hospitalized == 1)

###################################################
this_df = hosp_df %>% filter(covid==0)
results_0 = boot::boot(data=this_df,
                      statistic=burden_func_fast,
                      R=1000)
print(boot.ci(results_0, type="basic"))
###################################################
this_df = hosp_df %>% filter(covid==1)
results_1 = boot::boot(data=this_df,
                      statistic=burden_func_fast,
                      R=5000)
print(boot.ci(results_1, type="basic"))
```

#### nonhosp:
```{r, message=FALSE, warning=FALSE, echo=FALSE}
nohosp_df = burden_df %>%
  filter(hospitalized == 0)

###################################################
this_df = nohosp_df %>% filter(covid==0)
results_0 = boot::boot(data=this_df,
                      statistic=burden_func_fast,
                      R=1000)
print(boot.ci(results_0, type="basic"))
###################################################
this_df = nohosp_df %>% filter(covid==1)
results_1 = boot::boot(data=this_df,
                      statistic=burden_func_fast,
                      R=5000)
print(boot.ci(results_1, type="basic"))
```


### Table A. Demographic characteristics of patients without pre-existing COPD or Asthma by COVID status

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
ta_inc_df = cp_df %>%
  select(covid,
         age,
         clv_age,
         clv_race,
         clv_hisp,
         clv_sex)

ta_inc = ta_inc_df %>%
  tbl_summary(by = covid, missing = "no",
                statistic = list(
                  all_continuous() ~ "{mean} ({sd})",
                  all_categorical() ~ "{p}% ({n})"
                ),
                label = list(
                        clv_age ~ "Age groups",
                        clv_sex ~ "Sex",
                        clv_race ~ "Race",
                        clv_hisp ~ "Hispanic Ethnicity"
                      ),
                digits = list(
                  all_continuous() ~ c(1, 0),
                  all_categorical() ~ c(1, 0)
                )
              ) %>%
        add_p()

ta_inc
```

### Table B. Clinical characteristics of patients without pre-existing COPD or Asthma by COVID status 


```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
tb_inc_df = cp_df %>%
  select(covid,
         year_index,
         hospitalized, 
         criticalcare,
         ventilation,
         out_bins,
         em_bin,
         `emergency.visits.no.`,
         bmi_clv,     
         `Smoker..never`,
         `Smoker..current`,
         `Smoker..former`,
         `Smoker..missing`,
         death)

tb_inc = tb_inc_df %>%
  tbl_summary(by = covid,
                include = -c(criticalcare, ventilation),
                missing = "no",
                statistic = list(
                  all_continuous() ~ "{mean} ({sd})",
                  all_categorical() ~ "{p}% ({n})"
                ),
              digits = list(
                  all_continuous() ~ c(1, 0),
                  all_categorical() ~ c(1, 0)
                )
              ) %>%
        add_p()

tb_inc
```
```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
tb_inc_df = tb_inc_df %>% filter(hospitalized==1)

tb_inc = tb_inc_df %>%
  tbl_summary(by = covid,
                include = c(criticalcare, ventilation),
                missing = "no",
                statistic = list(
                  all_continuous() ~ "{mean} ({sd})",
                  all_categorical() ~ "{p}% ({n})"
                ),
              digits = list(
                  all_continuous() ~ c(1, 0),
                  all_categorical() ~ c(1, 0)
                )
              ) %>%
        add_p()

tb_inc
```

### Table C. Demographic characteristics of COVID+ patients with incident COPD or Asthma compared to COVID+ patients without evidence of condition 31 – 180 days post-COVID

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
tc_inc_df = cp_df %>%
  filter(covid==1) %>%
  select(flag_pulmonary,
         age,
         clv_age,
         clv_race,
         clv_hisp,
         clv_sex)

tc_inc = tc_inc_df %>%
  tbl_summary(by = flag_pulmonary, missing = "no",
                statistic = list(
                  all_continuous() ~ "{mean} ({sd})",
                  all_categorical() ~ "{p}% ({n})"
                ),
                label = list(
                      clv_age ~ "Age groups",
                      clv_sex ~ "Sex",
                      clv_race ~ "Race",
                      clv_hisp ~ "Hispanic Ethnicity"
                ),
              digits = list(
                  all_continuous() ~ c(1, 0),
                  all_categorical() ~ c(1, 0)
                )
              ) %>%
        add_p()

tc_inc
```

### Table D. Clinical characteristics of COVID+ patients with incident COPD or Asthma compared to COVID+ patients without evidence of condition 31 – 180 days post-COVID

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
td_inc_df = cp_df %>%
  filter(covid==1) %>%
  select(flag_pulmonary,
         year_index,
         hospitalized, 
         criticalcare,
         ventilation,
         out_bins,
         em_bin,
         `emergency.visits.no.`,
         bmi_clv,     
         `Smoker..never`,
         `Smoker..current`,
         `Smoker..former`,
         `Smoker..missing`,
         death)

td_inc = td_inc_df %>%
  tbl_summary(by = flag_pulmonary,
                include = -c(criticalcare, ventilation),
                missing = "no",
                statistic = list(
                  all_continuous() ~ "{mean} ({sd})",
                  all_categorical() ~ "{p}% ({n})"
                ),
              digits = list(
                  all_continuous() ~ c(1, 0),
                  all_categorical() ~ c(1, 0)
                )
              ) %>%
        add_p()

td_inc
```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
td_inc_df = td_inc_df %>% filter(hospitalized==1)

td_inc = td_inc_df %>%
  tbl_summary(by = flag_pulmonary,
                include = c(criticalcare, ventilation),
                missing = "no",
                statistic = list(
                  all_continuous() ~ "{mean} ({sd})",
                  all_categorical() ~ "{p}% ({n})"
                ),
              digits = list(
                  all_continuous() ~ c(1, 0),
                  all_categorical() ~ c(1, 0)
                )
              ) %>%
        add_p()

td_inc
```

### Table X. Incidence result components

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
cp_df = cp_df %>% mutate(
  inc_outcome_path = case_when(
    ###################################
    ((flag_pulmonary_asthma_dx == 1) | (flag_pulmonary_asthma_dx_med == 1))
          & ((flag_pulmonary_copd_dx == 0) & (flag_pulmonary_bronchiectasis_dx == 0)) ~ "Asthma is only diagnosis",
    ###################################
    ((flag_pulmonary_copd_dx == 1) | (flag_pulmonary_copd_dx_med == 1))
          & ((flag_pulmonary_asthma_dx == 0) & (flag_pulmonary_bronchiectasis_dx == 0)) ~ "COPD is only diagnosis",
    ###################################
    ((flag_pulmonary_bronchiectasis_dx == 1) | (flag_pulmonary_bronchiectasis_dx_med == 1))
          & ((flag_pulmonary_asthma_dx == 0) & (flag_pulmonary_copd_dx == 0)) ~ "Bronchiectasis is only diagnosis",
    ###################################
    (
      (flag_pulmonary_copd_dx == 1) | (flag_pulmonary_asthma_dx == 1) | (flag_pulmonary_bronchiectasis_dx == 1)
    ) ~ "Two or more dx among Asthma, COPD, or bronchiectasis",
    ###################################
    (flag_pulmonary_med == 1)
          & ((flag_pulmonary_copd_dx == 0) & (flag_pulmonary_bronchiectasis_dx == 0) & (flag_pulmonary_asthma_dx == 0)) ~ "Qualifying medication only (no dx in record)",
    ###################################
    T ~ "Qualifying medication only (no dx in record)" #<100 or 0.01%
  )
)

tx_inc_df = cp_df %>%
  select(
    covid,
    flag_pulmonary,
    inc_outcome_path,
    hospitalized
  )

tx_inc = tx_inc_df %>%
  tbl_summary(
    by=covid,
    include = c(flag_pulmonary),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{p}% ({n})"
    ),
              digits = list(
                  all_continuous() ~ c(1, 0),
                  all_categorical() ~ c(1, 0)
                )
  ) %>% add_p()

tx_inc
```


```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
tx_inc_df = tx_inc_df %>% filter(flag_pulmonary==1)

tx_inc = tx_inc_df %>%
  tbl_summary(
    by=covid,
    include = c(inc_outcome_path),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{p}% ({n})"
    ),
              digits = list(
                  all_continuous() ~ c(1, 0),
                  all_categorical() ~ c(1, 0)
                )
  ) %>% add_p()

tx_inc
```

### Upset plot of inclusion

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
list_input = list(
  any_asthma_DX = which(as.logical(cp_df$flag_pulmonary_asthma_dx == 1)),
  any_copd_DX = which(as.logical(cp_df$flag_pulmonary_copd_dx == 1)),
  any_bronchiectasis_DX = which(as.logical(cp_df$flag_pulmonary_bronchiectasis_dx == 1)),
  any_med = which(as.logical(cp_df$flag_pulmonary_med))
)

upset(fromList(list_input), nsets=90, order.by="freq")
```

### PPV Calculation

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
###############################################################
## Bring in redcap data, clean
raw_redcap_inc_df = read.csv("...", check.names = T)
colnames(raw_redcap_inc_df)[1] = "redcap_record_id"

filt_rerev_df = raw_redcap_inc_df %>% filter(
  !is.na(ssid_i) & ssid_i != "", # redcap ID isn't broken
  !is.na(dob_i) & dob_i != "", # you have a DOB with minimal quality
  "m" %in% sex_i | "M" %in% sex_i, # female/Female/male/Male
  incidence_complete == 2, # 2 means the reviewer said they did it fully
  index_lab_i %in% c(0,1) # covid y/n is populated
)

###############################################################
## Compute incident outcome
redcap_inc_df = filt_rerev_df %>% mutate(
  # baseline
  baseline_poor_airflow = case_when(
    as.numeric(pulm_inc_fev1) / as.numeric(pulm_inc_fvc) < 0.7 ~ T
  ),
  # followup
  followup_poor_airflow = case_when(
    as.numeric(pulm_postind_fev1_val) / as.numeric(pulm_postind_fev_val) < 0.7 ~ T
  )
) %>% mutate(
   # handle NA structure
   pulm_postind_copd = coalesce(pulm_postind_copd, 0),
   pulm_postind_rx = coalesce(pulm_postind_rx, 0),
   followup_poor_airflow = coalesce(followup_poor_airflow, 0),
   pulm_dx_prendex_i = coalesce(pulm_dx_prendex_i, 0),
   asthma_pre = coalesce(asthma_pre, 0),
   baseline_poor_airflow = coalesce(baseline_poor_airflow, 0),
) %>% select(
  ssid_i, redcap_record_id,
  pulm_postind_copd, pulm_postind_rx, followup_poor_airflow, pulm_dx_prendex_i, asthma_pre, baseline_poor_airflow
)

# redcap inc calc
redcap_inc_df$redcap_incident = (redcap_inc_df$pulm_postind_copd | redcap_inc_df$pulm_postind_rx | redcap_inc_df$followup_poor_airflow) & !(redcap_inc_df$pulm_dx_prendex_i | redcap_inc_df$asthma_pre | redcap_inc_df$baseline_poor_airflow)

###############################################################
## Compute PPV (redcap only given CP+ cases)
redcap_inc_df = redcap_inc_df %>% filter(!is.na(ssid_i) & ssid_i != "")

denom = redcap_inc_df %>% nrow()
tp = redcap_inc_df %>% filter(redcap_incident == T) %>% nrow()

print("PPV is # true positives / (false and true positive)")
print(tp/denom)
print("Sample size:")
print(denom)
```
 
#### Bootstrap PPV

```{r}

ppv_func = function(data, indices) {
  
  sample_data = data[indices, ] # resample the data
  denom = sample_data %>% nrow()
  tp = sample_data %>% filter(redcap_incident == T) %>% nrow()
  
  #print("PPV is # true positives / (false and true positive)")
  this_ppv = tp/denom
  return(this_ppv)
}
# bootstrapping with 1000 replications
results = boot::boot(data=redcap_inc_df,
                      statistic=ppv_func,
                      R=5000)

# view results
results
plot(results)

# get 95% confidence interval
boot.ci(results, type="basic")
```

### Description of PPV cohort

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
descrip_df = filt_rerev_df %>% select(
  sex_i,
  dob_i,
  index_lab_negative_date_i,
  index_lab_positive_date_i
)

descrip_df = descrip_df %>%
  mutate(
    dob_i = ymd(dob_i), # Convert to Date
    index_lab_negative_date_i = ymd(index_lab_negative_date_i), # Convert to Date
    index_lab_positive_date_i = ymd(index_lab_positive_date_i), # Convert to Date
    index_date = coalesce(index_lab_negative_date_i, index_lab_positive_date_i), # Pick non-NA value
    age_at_index = as.numeric(difftime(index_date, dob_i, units = "days")) / 365.25, # Calculate age
    sex_unified = case_when(
      sex_i == "Female" ~ "male",
      sex_i == "Male" ~ "male",
      sex_i == "female" ~ "female",
      sex_i == "female " ~ "female",
      sex_i == "FEMALE" ~ "female",
      sex_i == "M" ~ "male",
      sex_i == "male " ~ "male",
      sex_i == "male" ~ "male",
      sex_i == "F" ~ "female",      
      sex_i == "FEMALE " ~ "female",
      sex_i == "MALE " ~ "male",
      sex_i == "MALE" ~ "male",
      sex_i == "Female " ~ "female"
    ),
    age_brackets = case_when(
      age_at_index < 40 ~ "20-39 years",
      age_at_index < 55 ~ "40-54 years",
      age_at_index < 65 ~ "55-64 years",
      age_at_index < 75 ~ "65-74 years",
      age_at_index < 85 ~ "75-84 years",
      age_at_index >= 85 ~ "85+ years"
    )
  )

descrip_df %>% select(sex_unified, age_brackets) %>% tbl_summary()
```


## Tables for Worsening

```{r, include=F, eval=T}
# Load full CP
raw_cp_df_ex = read.csv("...", check.names=T)
raw_cp_df_ex$full_patid = paste(raw_cp_df_ex$site, "_", raw_cp_df_ex$patid, sep="")
raw_cp_df_ex = raw_cp_df_ex %>%
  filter(!site %in% c('miami', 'usf'))

raw_cp_df_ex = raw_cp_df_ex %>% mutate(
  ymd_index = ymd(index.date),
  year_index = case_when(
    year(ymd_index) %in% c(2020, 2021, 2022, 2023) ~ year(ymd_index)
  ),
  index_date = as.Date(index.date, format = "%Y-%m-%d")
) %>% filter(
  ymd_index <= as.Date("2023-02-01"),
  !is.na(year_index),
  age >= 20,
  index_date >= as.Date("2020-03-01"),
  !is.na(dob)
)
```


```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
cp_df = raw_cp_df_ex %>%
  mutate(year_index == factor(year_index))

cp_df = cp_df %>%
  mutate(clv_age = case_when(
    age < 40 ~ "20-39 years",
    age < 55 ~ "40-54 years",
    age < 65 ~ "55-64 years",
    age < 75 ~ "65-74 years",
    age < 85 ~ "75-84 years",
    age >= 85 ~ "85+ years"
  ))

cp_df = cp_df %>%
  mutate(clv_sex = case_when(
    Female == 1 ~ "Female",
    Male == 1 ~ "Male",
    T ~ "Other/Missing"
  ))

cp_df = cp_df %>%
  mutate(clv_race = case_when(
    Asian == 1 ~ "Asian",
    `Black.or.African.American` == 1 ~ "Black or African American",
    `White` == 1 ~ "White",
    `Other` == 1 ~ "Other",
    `Missing` == 1 ~ "Missing"
  ))

cp_df = cp_df %>%
  mutate(clv_hisp = case_when(
    `Hispanic..Yes` == 1 ~ "Hispanic: Yes",
    `Hispanic..No` == 1 ~ "Hipanic: No",
    `Hispanic..Other.Missing` == 1 ~ "Hispanic: Other/Missing"
  ))

cp_df = cp_df %>%
  mutate(out_bins = case_when(
    outpatient.no. == 0 ~ "0",
    outpatient.no. == 1 ~ "1",
    outpatient.no. < 6 ~ "2-5",
    outpatient.no. < 21 ~ "6-20",
    outpatient.no. > 20 ~ ">20"
  ))

cp_df = cp_df %>%
  mutate(em_bin = case_when(
    is.na(`emergency.visits.no.`) ~ "Missing",
    `emergency.visits.no.` > 1 ~ "2 or more",
    `emergency.visits.no.` == 1 ~ "1",
    `emergency.visits.no.` == 0 ~ "0"
  ))

cp_df = cp_df %>%
  mutate(bmi_clv = case_when(
    is.na(bmi) ~ "Missing",
    bmi < 18.5  ~ "Underweight (<18.5)",
    bmi < 25 ~ "Normal range (18.5 to <25)",
    bmi < 30 ~ "Overweight (25 to <30)",
    bmi >= 30  ~ "Obese (>30)"
  ))
```

### Excess burden bootstrap:

#### combined:

```{r, message=FALSE, warning=FALSE, echo=FALSE}

burden_func_fast_ex = function(data, indices, covid_bool) {
  sample_data = data[indices, ]
  percent = mean(sample_data$flag_worsening_pulmonary == 1) * 100
  
  return(percent)
}


burden_df = cp_df %>%
  select(hospitalized, covid, flag_worsening_pulmonary)

###################################################
this_df = burden_df %>% filter(covid==0)
results_0 = boot::boot(data=this_df,
                      statistic=burden_func_fast_ex,
                      R=1000)
print(boot.ci(results_0, type="basic"))
###################################################
this_df = burden_df %>% filter(covid==1)
results_1 = boot::boot(data=this_df,
                      statistic=burden_func_fast_ex,
                      R=5000)
print(boot.ci(results_1, type="basic"))
```

#### hosp:
```{r, message=FALSE, warning=FALSE, echo=FALSE}
hosp_df = burden_df %>%
  filter(hospitalized == 1)

###################################################
this_df = hosp_df %>% filter(covid==0)
results_0 = boot::boot(data=this_df,
                      statistic=burden_func_fast_ex,
                      R=1000)
print(boot.ci(results_0, type="basic"))
###################################################
this_df = hosp_df %>% filter(covid==1)
results_1 = boot::boot(data=this_df,
                      statistic=burden_func_fast_ex,
                      R=5000)
print(boot.ci(results_1, type="basic"))
```

#### nonhosp:
```{r, message=FALSE, warning=FALSE, echo=FALSE}
nohosp_df = burden_df %>%
  filter(hospitalized == 0)

###################################################
this_df = nohosp_df %>% filter(covid==0)
results_0 = boot::boot(data=this_df,
                      statistic=burden_func_fast_ex,
                      R=1000)
print(boot.ci(results_0, type="basic"))
###################################################
this_df = nohosp_df %>% filter(covid==1)
results_1 = boot::boot(data=this_df,
                      statistic=burden_func_fast_ex,
                      R=5000)
print(boot.ci(results_1, type="basic"))
```


### Table E. Demographic characteristics of patients with pre-existing COPD or Asthma by COVID status

```{r, message=FALSE, warning=FALSE, echo=FALSE}
ta_ex_df = cp_df %>%
  select(covid,
         age,
         clv_age,
         clv_race,
         clv_hisp,
         clv_sex)

#######################################################

ta_ex = ta_ex_df %>%
  tbl_summary(by = covid, missing = "no",
                statistic = list(
                  all_continuous() ~ "{mean} ({sd})",
                  all_categorical() ~ "{p}% ({n})"
                ),
                label = list(
                      clv_age ~ "Age groups",
                      clv_sex ~ "Sex",
                      clv_race ~ "Race",
                      clv_hisp ~ "Hispanic Ethnicity"
                ),
              digits = list(
                  all_continuous() ~ c(1, 0),
                  all_categorical() ~ c(1, 0)
                )
              ) %>%
        add_p()

ta_ex
```



### Table F. Clinical characteristics of patients with pre-existing COPD or Asthma by COVID status 

- table compiled from multiple tbl_summary components

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
cp_df = cp_df %>% mutate(
  baseline_path = case_when(
    ###################################
    ((flag_pulmonary_asthma_dx == 1) | (flag_pulmonary_asthma_dx_med == 1))
          & ((flag_pulmonary_copd_dx == 0) & (flag_pulmonary_bronchiectasis_dx == 0)) ~ "Asthma is only diagnosis",
    ###################################
    ((flag_pulmonary_copd_dx == 1) | (flag_pulmonary_copd_dx_med == 1))
          & ((flag_pulmonary_asthma_dx == 0) & (flag_pulmonary_bronchiectasis_dx == 0)) ~ "COPD is only diagnosis",
    ###################################
    ((flag_pulmonary_bronchiectasis_dx == 1) | (flag_pulmonary_bronchiectasis_dx_med == 1))
          & ((flag_pulmonary_asthma_dx == 0) & (flag_pulmonary_copd_dx == 0)) ~ "Bronchiectasis is only diagnosis",
    ###################################
    (
      (flag_pulmonary_copd_dx == 1) | (flag_pulmonary_asthma_dx == 1) | (flag_pulmonary_bronchiectasis_dx == 1)
    ) ~ "Two or more dx among Asthma, COPD, or bronchiectasis",
    ###################################
    (flag_pulmonary_med == 1)
          & ((flag_pulmonary_copd_dx == 0) & (flag_pulmonary_bronchiectasis_dx == 0) & (flag_pulmonary_asthma_dx == 0)) ~ "Qualifying medication only (no dx in record)",
    ###################################
    T ~ "Qualifying medication only (no dx in record)" #<100 or 0.01%
  )
)


tb_ex_df = cp_df %>%
  select(covid,
         baseline_path,
         year_index,
         hospitalized, 
         criticalcare,
         ventilation,
         out_bins,
         em_bin,
         `emergency.visits.no.`,
         bmi_clv,     
         `Smoker..never`,
         `Smoker..current`,
         `Smoker..former`,
         `Smoker..missing`,
         death)

tb_ex = tb_ex_df %>%
  tbl_summary(by = covid,
              include = -c(criticalcare, ventilation),
              missing = "no",
                statistic = list(
                  all_continuous() ~ "{mean} ({sd})",
                  all_categorical() ~ "{p}% ({n})"
                ),
              digits = list(
                  all_continuous() ~ c(1, 0),
                  all_categorical() ~ c(1, 0)
                )
              ) %>%
        add_p()

tb_ex
```


```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=F}
cp_df %>% filter(baseline_path == "Multiple criteria (test)") %>%
  select(flag_pulmonary_med,
         flag_pulmonary_copd_dx,
         flag_pulmonary_bronchiectasis_dx,
         flag_pulmonary_asthma_dx,
         flag_pulmonary_asthma_dx_med, # or these into the 1st 3 groups
         flag_pulmonary_copd_dx_med,
         flag_pulmonary_bronchiectasis_dx_med)
```


```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
tb_ex_df = tb_ex_df %>% filter(hospitalized==1)

tb_ex = tb_ex_df %>%
  tbl_summary(by = covid,
                include = c(criticalcare, ventilation),
                missing = "no",
                statistic = list(
                  all_continuous() ~ "{mean} ({sd})",
                  all_categorical() ~ "{p}% ({n})"
                ),
              digits = list(
                  all_continuous() ~ c(1, 0),
                  all_categorical() ~ c(1, 0)
                )
              ) %>%
        add_p()

tb_ex
```

### Table G. Demographic characteristics of COVID+ patients with pre-existing COPD or Asthma with exacerbation of condition versus no evidence of exacerbation 

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
tc_ex_df = cp_df %>%
  filter(covid==1) %>%
  select(flag_worsening_pulmonary,
         age,
         clv_age,
         clv_race,
         clv_hisp,
         clv_sex)

tc_ex = tc_ex_df %>%
  tbl_summary(by = flag_worsening_pulmonary, missing = "no",
                statistic = list(
                  all_continuous() ~ "{mean} ({sd})",
                  all_categorical() ~ "{p}% ({n})"
                ),
                label = list(
                      clv_age ~ "Age groups",
                      clv_sex ~ "Sex",
                      clv_race ~ "Race",
                      clv_hisp ~ "Hispanic Ethnicity"
                ),
              digits = list(
                  all_continuous() ~ c(1, 0),
                  all_categorical() ~ c(1, 0)
                )
              ) %>%
        add_p()

tc_ex
```

### Table H. Clinical characteristics of COVID+ patients with pre-existing COPD or Asthma with exacerbation of condition versus no evidence of exacerbation 

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
td_ex_df = cp_df %>%
  filter(covid==1) %>%
  select(flag_worsening_pulmonary,
         baseline_path,
         year_index,
         hospitalized, 
         criticalcare,
         ventilation,
         out_bins,
         em_bin,
         `emergency.visits.no.`,
         bmi_clv,     
         `Smoker..never`,
         `Smoker..current`,
         `Smoker..former`,
         `Smoker..missing`,
         death)

td_ex = td_ex_df %>%
  tbl_summary(by = flag_worsening_pulmonary,
              include = -c(criticalcare, ventilation),
              missing = "no",
                statistic = list(
                  all_continuous() ~ "{mean} ({sd})",
                  all_categorical() ~ "{p}% ({n})"
                ),
              digits = list(
                  all_continuous() ~ c(1, 0),
                  all_categorical() ~ c(1, 0)
                )
              ) %>%
        add_p()

td_ex
```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
td_ex_df = td_ex_df %>%
  filter(hospitalized==1)

td_ex = td_ex_df %>%
  tbl_summary(by = flag_worsening_pulmonary,
              include = c(criticalcare, ventilation),
              missing = "no",
                statistic = list(
                  all_continuous() ~ "{mean} ({sd})",
                  all_categorical() ~ "{p}% ({n})"
                ),
              digits = list(
                  all_continuous() ~ c(1, 0),
                  all_categorical() ~ c(1, 0)
                )
              ) %>%
        add_p()

td_ex
```

### Table I. Worsening results [Title TBD]

- This is distinct from the tables X/F/H above. This focuses on "how do you worsen?"
- Mutually exclusive split vs greedy split

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=F}
tx_ex_df = cp_df %>%
  select(
    covid,
    flag_worsening_pulmonary,
    flag_outcome_asthma_dx,
    flag_outcome_copd_dx,
    flag_outcome_infectious_pneumonia_dx,
    flag_outcome_respiratory_failure_dx,
    flag_outcome_med,
    hospitalized
  )

# Mutually exclusive split: 8onlydx, 9onlydx, (8|9|10|11dx), (12 medonly). Check 13/14 parallel to 8/9.

tx_ex_df = tx_ex_df %>% mutate(
  worsening_mode = case_when(
    ####################################
      ((flag_outcome_asthma_dx == 1))
        & ((flag_outcome_infectious_pneumonia_dx == 0) & (flag_outcome_respiratory_failure_dx == 0) & (flag_outcome_med == 0)) ~ "Asthma DX only",
    ####################################
      ((flag_outcome_copd_dx == 1))
        & ((flag_outcome_infectious_pneumonia_dx == 0) & (flag_outcome_respiratory_failure_dx == 0) & (flag_outcome_med == 0)) ~ "COPD DX only",
    ####################################
      ((flag_outcome_med == 1))
        & ((flag_outcome_infectious_pneumonia_dx == 0) & (flag_outcome_respiratory_failure_dx == 0) & (flag_outcome_asthma_dx == 0) & (flag_outcome_copd_dx == 0)) ~ "Medication only",
    ####################################
      ((flag_outcome_infectious_pneumonia_dx == 1) | (flag_outcome_respiratory_failure_dx == 1))
        & ((flag_outcome_med == 0) & (flag_outcome_asthma_dx == 0) & (flag_outcome_copd_dx == 0)) ~ "Other DX only",
    ####################################
      T ~ "Multiple criteria")
)

tx_ex = tx_ex_df %>%
  filter(hospitalized == 0) %>%
  tbl_summary(
    by=covid,
    include = c(flag_worsening_pulmonary),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{p}% ({n})"
    ),
              digits = list(
                  all_continuous() ~ c(1, 0),
                  all_categorical() ~ c(1, 0)
                )
  ) %>% add_p()

tx_ex
```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=F}
tx_ex_df = tx_ex_df %>% filter(flag_worsening_pulmonary==1)

tx_ex = tx_ex_df %>%
  tbl_summary(
    by=covid,
    include = c(worsening_mode),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{p}% ({n})"
    ),
              digits = list(
                  all_continuous() ~ c(1, 0),
                  all_categorical() ~ c(1, 0)
                )
  ) %>% add_p()

tx_ex
```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
tx_ex_df = cp_df %>%
  select(
    covid,
    flag_worsening_pulmonary,
    flag_outcome_asthma_dx,
    flag_outcome_copd_dx,
    flag_outcome_infectious_pneumonia_dx,
    flag_outcome_respiratory_failure_dx,
    flag_outcome_med,
    flag_outcome_asthma_med,
    flag_outcome_copd_med
  )

#  Greedy groups, top to bottom: 8onlydx, 9onlydx, (8|9|10|11dx), (12 medonly). Check 13/14 parallel to 8/9.

tx_ex_df = tx_ex_df %>% mutate(
  worsening_mode_greedy = case_when(
    ###################################
    ((flag_outcome_asthma_dx == 1))
          & (
             (flag_outcome_copd_dx == 0) &
             (flag_outcome_infectious_pneumonia_dx == 0) & 
             (flag_outcome_respiratory_failure_dx == 0) 
            ) ~ "Asthma is only diagnosis",
    ###################################
    ((flag_outcome_copd_dx == 1))
          & (
             (flag_outcome_asthma_dx == 0) &
             (flag_outcome_infectious_pneumonia_dx == 0) & 
             (flag_outcome_respiratory_failure_dx == 0) 
            ) ~ "COPD is only diagnosis",
    ###################################
    ((flag_outcome_infectious_pneumonia_dx == 1))
          & (
             (flag_outcome_asthma_dx == 0) &
             (flag_outcome_copd_dx == 0) & 
             (flag_outcome_respiratory_failure_dx == 0)
            ) ~ "infectious_pneumonia is only diagnosis",
    ###################################
    ((flag_outcome_respiratory_failure_dx == 1))
          & (
             (flag_outcome_asthma_dx == 0) &
             (flag_outcome_copd_dx == 0) & 
             (flag_outcome_infectious_pneumonia_dx == 0)
            ) ~ "respiratory_failure_dx is only diagnosis",
    ###################################
    (
      (flag_outcome_respiratory_failure_dx == 1) |
      (flag_outcome_asthma_dx == 1) |
      (flag_outcome_copd_dx == 1) | 
      (flag_outcome_infectious_pneumonia_dx == 1) 
    ) ~ "Two or more dx among Asthma, COPD, pneu, or resp_fail",
    ###################################
    (flag_outcome_med == 1) & 
      (
        (flag_outcome_respiratory_failure_dx == 0) &
        (flag_outcome_asthma_dx == 0) &
        (flag_outcome_copd_dx == 0) & 
        (flag_outcome_infectious_pneumonia_dx == 0) 
      ) ~ "Qualifying medication only (no dx in record)",
    ###################################
    T ~ "Qualifying medication only (no dx in record)" #<100 or 0.01%
  )
)

tx_ex = tx_ex_df %>%
  tbl_summary(
    by=covid,
    include = c(flag_worsening_pulmonary),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{p}% ({n})"
    ),
              digits = list(
                  all_continuous() ~ c(1, 0),
                  all_categorical() ~ c(1, 0)
                )
  ) %>% add_p()

tx_ex
```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
tx_ex_df = tx_ex_df %>% filter(flag_worsening_pulmonary==1)

tx_ex = tx_ex_df %>%
  tbl_summary(
    by=covid,
    include = c(worsening_mode_greedy),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{p}% ({n})"
    ),
              digits = list(
                  all_continuous() ~ c(1, 0),
                  all_categorical() ~ c(1, 0)
                )
  ) %>% add_p()

tx_ex
```

### Upset plot of inclusion

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}

list_input = list(
  any_asthma_DX = which(
    as.logical(cp_df$flag_outcome_asthma_dx == 1)),
  any_copd_DX = which(
    as.logical(cp_df$flag_outcome_copd_dx == 1)),
  any_pneumonia_DX = which(
    as.logical(cp_df$flag_outcome_infectious_pneumonia_dx == 1)),
  any_resp_failure_DX = which(
    as.logical(cp_df$flag_outcome_respiratory_failure_dx == 1)),
  any_med = which(
    as.logical(cp_df$flag_outcome_med))
)

upset(fromList(list_input), nsets=90, order.by="freq")
```

### PPV Calculation



```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
###############################################################
## Bring in redcap data
raw_redcap_ex_df = read.csv("...", check.names = T)
colnames(redcap_inc_df)[1] = "redcap_record_id"

# clean redcap inputs for quality
filt_rerev_df = raw_redcap_ex_df %>% filter(
  !is.na(ssid_e) & ssid_e != "",
  !is.na(dob_e) & dob_e != "",
  "m" %in% sex_e | "M" %in% sex_e,
  exacerbation_complete == 2,
  index_lab_e %in% c(0,1))

###############################################################
## Compute exacerbation outcome (LONG)

# Get sum (GINA score) across each row per 5 medication reviews
rerev_df = filt_rerev_df %>% mutate(
  asth_first_sum =  select(., starts_with("a1_med")) %>% rowSums(na.rm = TRUE),
  asth_second_sum = select(., starts_with("a2_med")) %>% rowSums(na.rm = TRUE),
  asth_third_sum =  select(., starts_with("a3_med")) %>% rowSums(na.rm = TRUE),
  asth_fourth_sum = select(., starts_with("a4_med_fu_e___")) %>% rowSums(na.rm = TRUE),
  asth_fifth_sum =  select(., starts_with("a4_med_fu_e_2")) %>% rowSums(na.rm = TRUE)
)
# Get maximum GINA count
rerev_df = rerev_df %>%
  rowwise() %>%
  mutate(
    max_meds = max(c(asth_first_sum, asth_second_sum, asth_third_sum, asth_fourth_sum, asth_fifth_sum))
  )
# Check if the max is in the first medication review, else exacerbation
rerev_df = rerev_df %>% 
  ungroup() %>%
  mutate(
    asthma_med_rev_exacerb = case_when(
      max_meds > asth_first_sum ~ 1,
      T ~ 0
    )
  )

# GOLD: like above but weighted. Issue with SAMA/biologic in the fields on redcap
# can we proceed with what we have? just go ahead ignoring SAMA (4 categories)
# focus on worsening to only (LABA), (LABA + LAMA), (LABA + LAMA + ICS)
# Up-weight
rerev_df$c2_med_fu_e___2 = 2*rerev_df$c2_med_fu_e___2
rerev_df$c2_med_fu_e___3 = 2*rerev_df$c2_med_fu_e___3
rerev_df$c2_med_fu_e___4 = 2*rerev_df$c2_med_fu_e___4
rerev_df$c3_med_fu_e___2 = 3*rerev_df$c3_med_fu_e___2
rerev_df$c3_med_fu_e___3 = 3*rerev_df$c3_med_fu_e___3
rerev_df$c3_med_fu_e___4 = 3*rerev_df$c3_med_fu_e___4
rerev_df$c4_med_fu_e___2 = 3*rerev_df$c4_med_fu_e___2
rerev_df$c4_med_fu_e___3 = 3*rerev_df$c4_med_fu_e___3
rerev_df$c4_med_fu_e___4 = 3*rerev_df$c4_med_fu_e___4

# Get sum (GOLD score) across each row per 5 medication reviews
rerev_df = rerev_df %>% mutate(
  copd_first_sum =  select(., starts_with("a1_med")) %>% rowSums(na.rm = TRUE),
  copd_second_sum = select(., starts_with("c2_med")) %>% rowSums(na.rm = TRUE),
  copd_third_sum =  select(., starts_with("c3_med")) %>% rowSums(na.rm = TRUE),
  copd_fourth_sum = select(., starts_with("c4_med_fu_e___")) %>% rowSums(na.rm = TRUE),
  copd_fifth_sum =  select(., starts_with("a4_med_fu_e_2")) %>% rowSums(na.rm = TRUE)
)
# Get maximum GOLD count
rerev_df = rerev_df %>%
  rowwise() %>%
  mutate(
    max_meds_copd = max(c(copd_first_sum, copd_second_sum, copd_third_sum, copd_fourth_sum, copd_fifth_sum))
  )
# Check if the max is in the first medication review, else exacerbation
rerev_df = rerev_df %>% 
  ungroup() %>%
  mutate(
    copd_med_rev_exacerb = case_when(
      max_meds_copd > copd_first_sum ~ 1,
      T ~ 0
    )
  )

# oneliner for new exacerb column:
# NA HANDLE CASES
rerev_df = rerev_df %>% mutate(
  ns_asthma_prior_e = coalesce(ns_asthma_prior_e, 0),
  ###
  pl_dx_base_e  = coalesce(pl_dx_base_e, 0),
  copd_dx_base  = coalesce(copd_dx_base, 0),
  copd_dx_base_2  = coalesce(copd_dx_base_2, 0),
  rx1_base  = coalesce(rx1_base, 0),
  ###
  a_ocs_e  = coalesce(a_ocs_e, 0),
  copd_e  = coalesce(copd_e, 0),
  pneum_e  = coalesce(pneum_e, 0),
  resp_e  = coalesce(resp_e, 0),
  c_ocs_e = coalesce(c_ocs_e, 0),
  asthma_e  = coalesce(asthma_e, 0),
  pneum_e_2  = coalesce(pneum_e_2, 0),
  resp_e_2  = coalesce(resp_e_2, 0),
  ###
  asthma_med_rev_exacerb  = coalesce(asthma_med_rev_exacerb, 0),
  copd_med_rev_exacerb  = coalesce(copd_med_rev_exacerb, 0)
)

rerev_df$new_exacerb =
  # exclusion
  !(rerev_df$ns_asthma_prior_e) &
  # baseline something
  (rerev_df$pl_dx_base_e | rerev_df$copd_dx_base | rerev_df$copd_dx_base_2 | rerev_df$rx1_base) &
  # followup condition worsening/med class worsening
  (rerev_df$a_ocs_e | rerev_df$copd_e | rerev_df$pneum_e | rerev_df$resp_e | rerev_df$c_ocs_e | rerev_df$asthma_e | rerev_df$pneum_e_2 | rerev_df$resp_e_2 |
     rerev_df$asthma_med_rev_exacerb | rerev_df$copd_med_rev_exacerb)
rerev_df = rerev_df %>% select(ssid_e, new_exacerb)

###############################################################
## Compute PPV (redcap only given CP+ cases)

denom = rerev_df %>% nrow()
tp = rerev_df %>% filter(new_exacerb == T) %>% nrow()

print("PPV is # true positives / (false and true positive)")
print(tp/denom)

print("Sample size:")
print(denom)
```

#### PPV bootstrap

```{r}

ppv_func_inc = function(data, indices) {
  
  sample_data = data[indices, ] # resample the data
  denom = sample_data %>% nrow()
  tp = sample_data %>% filter(new_exacerb == T) %>% nrow()
  
  #print("PPV is # true positives / (false and true positive)")
  this_ppv = tp/denom
  return(this_ppv)
}
results = boot::boot(data=rerev_df,
                      statistic=ppv_func_inc,
                      R=5000)

# view results
results
plot(results)

# get 95% confidence interval
boot.ci(results, type="basic")
```

### Description of PPV cohort

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=T}
descrip_df = filt_rerev_df %>% select(
  sex_e,
  dob_e,
  index_lab_negative_date_e,
  index_lab_positive_date_e
)

descrip_df = descrip_df %>%
  mutate(
    dob_e = ymd(dob_e), # Convert to Date
    index_lab_negative_date_e = ymd(index_lab_negative_date_e), # Convert to Date
    index_lab_positive_date_e = ymd(index_lab_positive_date_e), # Convert to Date
    index_date = coalesce(index_lab_negative_date_e, index_lab_positive_date_e), # Pick non-NA value
    age_at_index = as.numeric(difftime(index_date, dob_e, units = "days")) / 365.25, # Calculate age
    sex_unified = case_when(
      sex_e == "Female" ~ "male",
      sex_e == "Male" ~ "male",
      sex_e == "female" ~ "female",
      sex_e == "FEMALE" ~ "female",
      sex_e == "M" ~ "male",
      sex_e == "male " ~ "male",
      sex_e == "male" ~ "male",
      sex_e == "F" ~ "female",      
      sex_e == "FEMALE " ~ "female",
      sex_e == "MALE " ~ "male"
    ),
    age_brackets = case_when(
      age_at_index < 40 ~ "20-39 years",
      age_at_index < 55 ~ "40-54 years",
      age_at_index < 65 ~ "55-64 years",
      age_at_index < 75 ~ "65-74 years",
      age_at_index < 85 ~ "75-84 years",
      age_at_index >= 85 ~ "85+ years"
    )
  )

descrip_df %>% select(sex_unified, age_brackets) %>% tbl_summary()
```

### Plot fig 2

- from .xlsx input

```{r}
df = read_xlsx('...')
df = df %>% mutate(
  label = paste(outcome, strata, covid, sep = " "),
  covid = recode(covid,
                 "pos" = "(+)",
                 "neg" = "(-)"),
  strata = recode(strata,
                  "nh" = "Not hospitalized",
                  "hosp" = "Hospitalized",
                  "all" = "All patients")
)


# Plot with custom colors
inc_df = df %>% filter(outcome == "inc")
ex_df = df %>% filter(outcome == "ex")

ggplot(inc_df, aes(x = strata, y = point, ymin = lower, ymax = upper, fill = covid)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.6) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    position = position_dodge(width = 0.8), # Match this to geom_bar
    width = 0.38,
    linewidth = 0.8,
    linetype = "solid",
    alpha=0.4) +
  geom_text(
    aes(label = round(point, 1)),
    position = position_dodge(width = 0.8),
    hjust = 1.6,  # Adjusts text position to the right of the bar
    size = 3.5
  ) +
  coord_flip() +
  scale_fill_manual(values = c("(-)" = "lightblue", "(+)" = "orange")) +
  scale_y_continuous(breaks = seq(0, max(df$upper)+2, by = 2)) +
  labs(
    title = "Incidence",
    x = "",
    y = "Per 100",
    fill = "COVID-19 Status"
  ) +
  theme_minimal()

```




